DynaProg for Scala

### 2013-01-01 docs/roadmap/design.tex
More report, fixed related nits.

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Recurrences analysis}
Since the original ADP introduces many concatenation combinators to differentiate between empty/non-empty, and floating/fixed-length concatenations, it might become quite involved for the programmer to make sure that the concatenation operator exactly matches the size of each subsequence involved. Additionally the priority of operators might vary in Scala, depending the operator's first character whereas it is possible to define arbitrary priorities in Haskell. To overcome these problems, we propose to automate the computation of the length of the left and right handside subwords of concat operators.

Let us introduce the concept of dependency: a dependency between tabulations $A,B$ denoted $A\to B$ exists if $B_{(i,j)} = f(A_{(i,j)})$, that is if the result of $B$ depends of the result on the \textit{same} subproblem computed in $A$. A grammar is unsolvable if there exist a dependency loop between parsers ($A\to ... \to A$). Such case only happen when there is no concatenation or a concatenation with an empty word.

{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}

Knowing that fact we can now easily track the dependencies of tabulations and infer an order of computation between them. Although unnecessary in a top-down approach (Scala), this order becomes essential in a bottom-up approach (CUDA) to make sure we always have all dependencies computed. Once we have such an order we can also trigger a min/max analysis on parsers length. Since a parser is made of terminals, concat, operations (aggregate, map, filter) and tabulations and since the min/max length of terminals is fixed, we only need to assign appropriate size to tabulations,  other operation simply need to propagate the information.

It is possible to obtain the yield size using the following algorithm (assuming all parser loops contain at least one tabulation):\ol
\item Set the yield minimum size of all tabulations to a large
\item Repeat $k$ times ($k$ is the number of tabulations): for each tabulation, compute its minimum yield size and update its value (without recursion at tabulations). This would lead to a correct minimum yield size because the terminals provide a minimum size and this might need at most $k$ iteration to propagate across all tabulations.
\item Set the maximal yield of all the tabulations to the minimal value. For each tabulation, compute recursively up to depth $k$ (where the depth is computed as the number of tabulation traversed) the maximum yield size. If the depth reaches the maximum $k$, there is a loop between tabulations, hence return infinity.
\ole

The last part of this algorithm seems to have an exponential complexity, however, if we consider depth-first search and return as soon as we reach infinity, we might reduce its complexity to $O(k^2)$. Obtaining the yield size of tabulations provides the following benefits:\ul
\item Minimum size: prevents self-loops (computing one matrix cell using its result) and avoid considering subsequences that would yield empty result (hence reducing time complexity).
\item Maximum size: allows to reduce the size of the result and backtrack matrix to $O(k\cdot n)$ instead of $O(n^2)$ (where $k$ is the maximum yield size), possibly providing substantial space savings.
\ule

% ------------------------------------------------------------------------------------------------
\subsection{Backtracking}
In order to do a clean and efficient transformation from ADP-like language to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules. To do that, we slightly need to modify the ADP language in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack writes are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial). To deal with this problem, we are facing two options:\ul
\item \textbf{Explicit backtracking:} requires clear syntactical separation between the score and the backtrack and is not implemented in vanilla ADP (unless the whole backtrack is made part of the scoring: big performance impact and non-constant memory requirement issues make its GPU implementation hard and not desirable). Since the backtracking is user-defined, there is no way to generate the backtracking algorithm automatically, hence the user also needs to provide it. 
\item \textbf{Implicit backtracking:} implies that every rule needs to be normalized, and transformed such that given a rule identifier and a set of indices (subproblems breaking), it is possible to retrieve the subproblems combination that build the problem. To do that we need to apply the following transformations\ol
	\item Normalize rules and identify them uniquely by \ul
		\item Removing all the <<or>> operators by breaking a parser into multiple subrules
		\item Forwarding names: replace a rule that is a single invocation by its content (i.e. name a terminal parser with its own name instead of its containing parser's)
		\item Assigning each subrule an unique identifier $r$ and create a mapping table $T: r \to$ user-defined name of the parser.
	\ule
	\item The data elements corresponding to a rule are (scoring, backtracking) and named after the tabulation. The scoring is a user-defined primary type (int, double, ...) and the backtracking is a tuple $({\rm id}_{\rm subrule}, k_1, k_2, \ldots , k_m)$ where $m$ is the maximal number of splits that can occur in the subrules.
	\item During the matrix computation of cell $(i,j)$, if the subrule $r$ applies, the backtrack will be set as $(r,k_1,k_2,\ldots,k_{m_r})$ (obviously with $i\le k_1\le k_2\le \ldots k_{m_r}\le j$ and $m_r\le m$).
	\item During backtracking, when reading the cell $(i,j)$ with backtrack $(r,k_1,k_2,\ldots,k_m)$, given $r$, we recover the subrule hence the user-defined name of the rule that applies and $k_{m_r}$, which allows us to enqueue the subwords $(i,k_1), (k_1,k_2), ..., (k_{m_r})$ for further backtracking. If $r$ refers to a terminal, we stop the backtracking.
	\item The backtracking can be returned to the user as a mapping table $T$ and a list of triplets $(r,i,j)$ where $r$ is the index of the sub-rule that can be easily mapped into the user name (such that $T: r_u \to$ user-defined rule name) and $(i,j)$ is the subword on which the rules applies.
\ole
In short, we break parsers into normalized rules, the backtracking information is the rule id (which rule to unfold) and a list of indices (how to unfold it).
\ule
Assuming that the backtracking information is meant to guide further processing, we provide this information into a list constructed bottom-up: it can be simply processed by the user program in-order, applying for each user-defined rule the underlying transformation, and storing intermediate results (i.e. in a hash map) until they are processed by another rule. Since consumed sub-result can be dropped and since the backtracking contains only relevant sub-problems, ultimately, only the result will be stored.

% ----------------------------------------------
\subsubsection{Backtracking with multiple backtrack elements}
The backtracking technique described above work fine when there is a single element stored per matrix cell (which is usually the case with min/max problems). However, in the generalization introduced by ADP, it is possible that a matrix cell stores multiple results. In such case, we need to select a correct result to avoid backtracking inconsistencies.

>>>>

% ------------------------------------------------------------------------------------------------
\newpage
\subsection{Recurrences analysis}
\subsubsection{Dead rules elimination}
Dead rules\footnote{\textit{Rule} denotes a tabulation belonging to the grammar; both terms refer to the same concept, with the subtle difference that \textit{rule} emphasizes its grammar membership.} elimination analysis is straightforward: starting from the grammar's axiom, recursively collect all tabulations involved in the computation in $R$ (set of rules that are mandatory in the grammar). The dead rules $D=S \setminus R$ (where $S$ is the set of all tabulations) can safely be removed from the grammar rules. Although seemingly useless for the Scala implementation, this step is necessary to maintain coherency between Scala and CUDA rules numbering (that happen in a later stage on the valid rules). In CUDA, this analysis not only provides dead code elimination, but it also prevents useless computation execution, since all rules are computed sequentially for a particular subsequence before the next subsequence is processed.

\subsubsection{Yield analysis}
Since the original ADP introduces many concatenation combinators\footnote{ADP's original combinators are: $\sim\sim, \sim\sim*, *\sim\sim, *\sim*, -\sim\sim, \sim\sim-, +\sim\sim, \sim\sim+, +\sim+$.} to differentiate empty/non-empty, and floating/fixed-length concatenations, it is quite involved for the programmer to make sure that the concatenation operators exactly match the size of each pair of subsequence involved. Additionally the priority of operators varies in Scala, depending the operator's first character whereas it is possible to specify arbitrary priorities in Haskell. To overcome these issues, we propose to automate the computation of the minimum/maximum length of (subsequences) parsers.

Parsers are made of terminals, concat, operations (aggregate, map, filter) and tabulations; minimum/maximum yield of terminals is set, hence it only remains to assign appropriate yield sizes to tabulations; other operations simply propagate that information. It is possible to obtain the yield size of tabulations using the following algorithm (assuming recursive parsers contain at least one tabulation that is part of the loop):\ol
\item Set the yield minimum size of all tabulations to a large number $M_0$ (such that all tabulation would reasonably have a minimum yield size smaller than $M_0$)
\item Repeat $k$ times ($k$ is the number of rules of the grammar): for each rule, compute its minimum yield size and update its value (without recursion at tabulations). This would lead to a correct minimum yield size because the terminals provide a minimum size and this might need at most $k$ iteration to propagate across all rules.
\item Set the maximal yield of all the rules to the minimal value. For each rule, compute recursively up to depth $k$ (where the depth is computed as the number of tabulation traversed) the maximum yield size. If the depth reaches the maximum $k$, there is a loop between tabulations, hence return infinity.
\ole

The last part of this algorithm seems to have an exponential complexity, however, if we consider depth-first search and return as soon as we reach infinity, we might reduce its complexity to $O(k^2)$. Obtaining the yield size of tabulations provides the following benefits:\ul
\item Minimum size: prevents self-loops (computing one matrix cell using its own result) and avoid considering subsequences that would yield empty result (hence slightly reducing time complexity).
\item Maximum size: allows to reduce the size of the result and backtrack matrix to $O(m\cdot n)$ instead of $O(n^2)$ (where $m$ is the maximum yield size), possibly providing substantial space savings. As the rules with bounded maximum yield are very rare, we did not implement this optimization, although we might consider it for future work.
\ule

\subsection{Dependency analysis}
Let us introduce the concept of dependency: a dependency between tabulations $A,B$ denoted $A\to B$ exists if $B_{(i,j)} = f(A_{(i,j)})$, that is if the result of $B$ depends of the result on the \textit{same} subproblem computed in $A$. A grammar is unsolvable if there exist a dependency loop between parsers ($A\to ... \to A$). Such case only happen when there is no concatenation or a concatenation with an empty word. Being able to track the dependencies of tabulations and infer a computation order between them has two benefits:\ul
\item Although seemingly unnecessary in a top-down approach (Scala), this analysis detects dependency loops which would result in infinite call loops (stack overflow) at execution.
\item Ordering tabulations is critical in a bottom-up approach (CUDA) to make sure that all dependencies are valid before an element computation is triggered.
\ule

% ------------------------------------------------------------------------------------------------
\subsection{Backtracking}
In order to produce an efficient transformation from ADP-like problem description to plain C recurrences, we need to construct bottom-up recurrences from top-down parser rules. To do that, we slightly need to modify the ADP parsers in order to separate the backtracking and the scoring, because we want to obtain an efficient algorithm: backtrack writes are in $O(n^2)$ whereas score reads are proportional to the algorithmic complexity ($O(n^3)$ or more for non-serial). To deal with this problem, we are facing two options:\ul
\item \textbf{Explicit backtracking:} requires clear syntactical separation between the score and the backtrack which is not implemented in ADP, unless we consider the whole backtrack being part of the scoring (which has a big performance impact and non-constant memory requirement issues that make such GPU implementation hard and not desirable). Additionally, since the backtracking data is user-defined, there is no way to generate the backtracking algorithm automatically, hence the user also needs to provide it.
\item \textbf{Implicit backtracking:} implies that every rule needs to be normalized, and transformed such that given a rule identifier and a set of indices (subproblems breaking), it is possible to retrieve the sub-solutions combination that contribute to the problem solution. To do that we need to apply the following transformations\ol
	\item Normalize rules and identify them uniquely by exploding alternatives: each rule is decomposed into the union of multiple sub-rules uniquely identified by an index, where sub-rules do not contain alternatives (Or parsers). Let $s$ a subrule and $r_s$ its identifier, we also establish a mapping $T$ from identifier to subrule: $(r_s \to s) \in T$.
	\item Let ${\rm cc}(r_s)$ the number of concatenation contained in the sub-rule $r_s$. The data element corresponding to a rule is a pair (score, backtrack) and is named after the tabulation.\ul
		\item The score part consists of a user-defined type (a composite of primitive types, case classes and tuples)
		\item The backtrack part is a tuple $(r_s,(k_1, k_2, \ldots , k_m))$ where $m$ is the maximal number of concatenations occurring in the enclosing rule of $r_s$; more formally $m=\max_z\big[{\rm cc}(r_z) | r_z \in {\rm rule}(r_s)\big]$, and let $m_s = {\rm cc}(r_s) \le m$.
	\ule
	\item During the matrix computation of cell $(i,j)$, if the sub-rule $r_s$ applies, the backtrack will be set as $(r_s,(k_1,k_2,\ldots,k_{m_s}))$; with $i\le k_1\le k_2\le \ldots k_{m_s}\le j$. Note that if the backtrack occupies a fixed-length memory, the backtrack will contain exactly $m$ indices, hence $k_i\,|\,m_s<i\le m$ will be unspecified.
	\item During backtracking, when reading the cell $(i,j)$ with backtrack $(r_s,(k_1,k_2,\ldots,k_m))$, given $r_s$, we recover $s=T(r_s)$, the sub-rule that applies. Hence we can determine $m_s$, which allows us to enqueue the subsequences $(i,k_1), (k_1,k_2), ..., (k_{m_s},j)$ for recursive backtracking. If $s$ refers to a terminal, we stop the backtracking.
	\item The backtracking can be returned to the user as a mapping table $T$ and a list of triplets $((i,j),r_s,(k_1,k_2,\ldots,k_{m_s}))$ where $(i,j)$ denotes the subsequence on which the sub-rule $T(r_s)$ has to be applied with concatenation indices $(k_1,k_2,\ldots,k_{m_s})$.
\ole
In short, we break parsers into normalized rules, the backtracking information is the subword, the sub-rule id (which rule to unfold) and a list of indices (how to unfold it).
\ule
In order to reduce the storage required by the backtracking indices, we can avoid storing fixed indices (where at least one of the two subsequences involved in a concatenation has a fixed size) and leverage the knowledge contained in $s$ to reconstruct the appropriate backtrack.

Assuming that the backtracking information is meant to guide further processing, we provide this information into a list constructed bottom-up: it can be simply processed in-order, applying for each rule the underlying transformation, and storing intermediate results (i.e. in a hash map) until they are processed by another rule. Since there is only one consumer for each intermediate result, every read in the hash map can remove the corresponding value, hence reducing the memory consumption. Ultimately, only the problem solution will be stored in the hash map.

% ----------------------------------------------
\subsubsection{Backtracking with multiple backtrack elements}
{\center\color{red} XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX\\ CONTINUE HERE\\ XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX}

The backtracking technique described above work fine when there is a single element stored per matrix cell (which is usually the case with min/max problems). However, in the generalization introduced by ADP, it is possible that a matrix cell stores multiple results. In such case, we need to select a correct result to avoid backtracking inconsistencies.
